---
title: "Introduction to Urban Economics"
subtitle: "EC 330, Set 01"
author: "Andrew Dickinson"
date: "Fall 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'nord.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# knit: pagedown::chrome_print
---
class: inverse, center, white

```{R, setup, include = F}
options(htmltools.dir.version = FALSE)
# Load packages
 pacman::p_load( 
      broom, here, tidyverse, latex2exp, ggplot2, ggthemes, viridis, extrafont, magrittr,
      gridExtra, kableExtra, dplyr, magrittr, knitr, emo,  hrbrthemes, sf, tidycensus) 

# Define colors
## source color script from local directory
## comment out and replace colors if user is not Andrew
source("~/Documents/scripts/colors/colors.R")
# XaringanExtra functions
## Tile view
xaringanExtra::use_tile_view()
## Scribble
xaringanExtra::use_scribble()
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})


```


```{r, echo = FALSE, cache = FALSE, results = FALSE, message = FALSE, include = FALSE, error = FALSE}
income <-  tidycensus::get_acs(
 geography = "county", variables = "B19013_001",
 state = c("WA", "OR", "CA"), geometry = TRUE, cache = TRUE
)
``` 

```{css, echo = F, eval = F}
@media print {
  .has-continuation {
    display: block !important;
  }
}
```

class: inverse, middle
# Prologue

---
# Schedule

.bigger[.hi[Today]]
- Welcome, check in, survey, and syllabus
- Introduction: .it[What is (urban) economics?]
- Toolkit: .it[What is a model?]

--

.bigger[.hi[Upcoming]]
- EC201 review
- Principals of urban economics


---
#Introduction: About me

.hi[.ul[Admin:]] Just call me .h[Andrew]

- Office: [PLC 523](https://map.uoregon.edu/7c6d0ffbd), Office Hours: .hi[TBD], Email: .hii[adickin3@uoregon.edu]


--

.hi-blue[.ul[School:]]

- .hi[Fourth year] PhD student researching:
  - Applied micro topics related to environmental economics
  - Causal inference, ML, and data science

--

.hi-green[.ul[Not school:]]

  - Grew up in San Diego, CA
  - I enjoy spending time outside and listening to good music `r emo::ji("call me hand")`
  
---
# Introduction: About you

.bigger[.h[I hope that you:]]

--

.hi[(i)]   Are an eager student ready to learn about urban economics

--

.hi[(ii)]  Have passed EC201 with at least .hi-blue[some recollection] of the material

--

.hi[(iii)] Engage in class (ask lots of questions) and support your peers

--

.hi[(iv)]  Will read my [syllabus](https://www.youtube.com/watch?v=Tnlaokj1opA) .smallerer[(especially before sending me an email)]

<br>

--

.bigger[.h[I want to discuss the following:]]

--

.pull-left[
.hi[(i).] Course policies
]
.pull-right[
.hi[(ii).] Grades and grading]

---
class: inverse, middle

# Syllabus

---
# Syllabus: Attendance

Attendance is ".hi[mandatory]", though
- it does .hi-red[not] impact your grade
- records will .hi-red[not] be kept (though I have a good memory)
- lectures will .hi-red[not] be recorded

--

Attendance is an efficient use of your time
- Mixture of .hi-blue[slide decks] and .hi-green[in-class examples] on the board

--

.hi[Slide decks] will be posted to the course page .smaller[(.l[https://ajdickinson/EC330F22])] _sometime_ after lecture<sup>.hii[†]</sup>

.footnote[<sup>.hii[†]</sup> This is the only course material that will not be posted to the Canvas page. Examples will not be posted]

--

Exams will be administered during class time

---
# Syllabus: Grading

There are .hi-red[500] points total.  .hi[Your grade will be determined by:]

--

- 45%: Final Exam: .hi[225] pts

- 30%: Midterm Exam: .hi[150] pts

- 25%: Problem sets: 25 pts each, .hi[75] pts total

- 10%: Book Report: .hi[50] pts

--

<br>

.center[.hi[Q]: Is there a curve?]

--

.center[.hi[A]: Typically yes. At the end of the term the class will be curved ]

---
# Syllabus: Grading

.hi-blue[Caveat]: Due to a curve, standing in this course will depend on your score relative to your peers--I may not be able to give you an accurate assessment of your grade

--

From the syllabus:

> Your grade will be determined relative to your peers, so during the course, I will not be able to tell you what your exact letter grade is at any point in time, because it depends on everyone’s overall scores of the class.

--

<br>

Following department policies, for 300 and 400 level classes roughly 65% of the class will receive A’s and B’s.

---
# Syllabus: Exams

.hi[Midterm:] Wed, Nov 2 (in class)

.hi[Final:] [Wed, Dec 7 @ 14:45](https://registrar.uoregon.edu/calendars/examinations#ay22)

--

.center[.hi[Absolutely no makeups]]

<br>
  
--

Under .hi[extraordinary circumstances] I will shift midterm weight to the final
- Entirely by my discretion
- Must contact me before the exam via email
    
--

<br>

.hi[Drop this course if you are unable to take the scheduled midterm]


---

#Syllabus: TotC

Required reading [Triumph of the City](https://www.amazon.com/Triumph-City-Greatest-Invention-Healthier/dp/0143120549) by Ed. Glaeser. 

Easy reading that is .hi[suuuuper interesting], covering many of the same topics that we will in class; building your intuition

--

.hi[Book Report:] Due Sunday following the last day of class (Dec 4 @ 11:59p)
- Instructions and a rubric are posted on the course page + canvas soon
- Straight forward, easy points--so long as you read

--

.hi-red[Midterm] and .hi-red[final] will have questions connecting key topics in the books to topics in lecture
- If you read they should be easy; if you do not read they will be painful

---
#Syllabus: Problem sets

There will be .hi[4] problem sets throughout the term:
- Signal what to expect on the exams
- You must submit a .hi[.pdf] document on Canvas
- The lowest scored problem set is dropped
- No late homework assignments will be accepted
- If you scan is bad, or your handwriting illegible, you get .hi-red[nothing]

--

The problem sets are your .hi[best study tool] for the .hi-red[exams]. Do all of them!

--

.hi[General tips:]

--

- Start them .hi[early]
- Bring your work to class and ask for help
- Go to my office hour

---
# Syllabus: Canvas + course site

I use Canvas in conjunction to a simple course website hosted on GitHub

--

.bigger[.hi[Canvas:]]
- .hi[All] assignments submissions and announcements

--

<br>

.bigger[[Course site:](https://github.com/ajdickinson/EC330F22)]
- Light, public page with links to all course material
- Do not need an account
- I use it out of convenience and other personal reasons
- .hi[ONLY] thing on the page that will not be on Canvas is the slide decks

---
class: inverse, middle, center
# What is economics?

---
# What is economics?

.hi[_The Wealth of Nations_] _by_ [Adam Smith](https://www.adamsmith.org/about-adam-smith) (1776)

--

Seminal work on economics, four volumes; comparable to the works of:
- Newton’s _Principia Mathematica_
- Darwin’s _Origin of Species_
  
--

.hi[Wealth] ("weal")

--

- Money and other assets
- But also well-being, welfare
  
--
  
.hi[Archaic] use of "wealth" 
- Economics is not merely the study of how to get rich
- Please don't ask me about the stock market

---
# What is economics?

In a nutshell:

.center[.hi[Economics is the study of how people make allocation decisions to maximize their happiness when facing limited resources (budgets)]]

--

There is never .hi[enough resources]; we call this .hi-red[scarcity]

--

.hi-red[Scarcity] gives us constraints to which we do the best _we can_

--

.hi[Other examples of scarcity]
.pull-left[
  - Time
- Health]
.pull-right[
  - Land
- Natural resources]

--

.hi[Economics is very general;] the market lens can be used across many topics


<!-- --- -->
<!-- # Economic perspective -->

<!-- Economics uses a positive perspective on the world -->
<!-- - about how actors .hi[actually] behave; less about how they .hi[should] behave -->

<!-- -- -->

<!-- .hi-blue[Normative statements:] Matters of opinion; may never be resolved -->
<!-- - Higher ed should be free -->

<!-- .hi[Positive statements:] Matters of fact that can be resolved by data -->
<!-- - Who should get federal grants? The best students or students in need? -->

<!-- -- -->

<!-- Figure out what the data say for evidence-based policy making -->

<!-- -- -->

<!-- .hi[Opportunity cost:] What you have to give up by not putting a resource to its _next-best_ alternative use -->

<!-- - Economist use the word .hi[cost] differently than the general public -->
<!--   - "Cost" is different than "price" -->

---
class: inverse, center, middle
# Intro to Urban Economics

---
name:what_is_it
# Intro to Urban Economics

A mash-up between .hi[economics] and .hii[geography] 

--

.hi[Economics:] Study of how people and firms allocate scarce resources.
- Individual utility, markets, trade, welfare

--

.hii[Geography:] Studies effects of location and the environment 
- Hydrology, climate, resources, etc.

--

<br>

.hi[Economics] + .hii[Geography]  : Study of how individuals and firms choose utility and profit maximizing locations, and consequences of these decisions

---
# Intro to Urban Economics

We will study how the .hi[distribution] of people & firms across space impacts:

--

.pull-left[
  - City growth `r emo::ji("cityscape")`
- Crime `r emo::ji("policewoman")`
- The environment `r emo::ji("recycle")`]
.pull-right[
  - Income growth & inequality  `r emo::ji("chart increasing")`
- Education `r emo::ji("school")`
- Employment + wages `r emo::ji("money-mouth face")`]
  
--

We will also examine the efficacy of various .hi[place - based policies]
.pull-left[
  - Minimum Wage
- Rent Control]
.pull-right[
  - Land Use Restrictions
- Sustainability]

--

<br>

.hi[What are the economic drivers behind urban development?]

---
# This Course

This class has two fairly distinct halves:

1. .hi[Philosophy & Tools]

  - Why do cities exist? Why do they grow? Why do they decline?
  
  - Fundamental tools of labor & urban econ (supply and demand)
  
--

2. .hi[Application]

  - Rent Control & Minimum Wage
  
  - Highways and urban transportation
  
  - Income inequality and environmental issues

---
# What is a city?

.hi[According to the Census Bureau...]

--

.hi[Urban Area:] a densely settled geographical area with:

--

  - Minimum population of .hi[2,500]
  - Minimum density of .hi[500 people per square mile]

--

.hi[Metropolitan Area:] an urbanized area with at - least .hi[50k population]

--

.hi[Micropolitan Area:] an urbanized area with at least .hi[10k] but not as many as .hi[50k] people

--

.hi[MSA:] abbrev. for .slate[both] .hi[metropolitan] and .hi[micropolitan] statistical area

--

.hi[Principal City:] the .hi-blue[largest municipality in an MSA]

--

.hi[City:] .blue[Dense collection of people in specific geographic area]

---
# Intro to Urban: Cities

The majority of the US population lives in cities<sup>.smallest[.hi-pink[†]]</sup> and more people will continue to migrate into urban areas

.footnote[.hi-pink[†]. 80 - ish percent, according to the Census Bureau]

.hi[Questions:]

--

1. Do you like cities?

2. What are favorite city .hii[amenities]?

3. Are cities at odds with the natural world?

--

.hi[Why do cities exist?]

--

.center[.hi[Location matters]]


---
# Intro to Urban: Location...

Where you live has implications for

--

- Your contribution to .hi[global carbon emissions] 

  - _Why does this vary across cities?_

--

- Your .hi[wage] and .hi[rent] 

  - _Why does this vary across cities?_

--

- Your .hii[economic mobility]

  - _Why does this vary across cities?_
  
--

.hi[We will answer these questions as we progress through this course]

---
name:what_is_model
# What is a model?

In this class we will make use of various .hi[mathematical models]

--

What is a .hi[mathematical model]?
  - A model is a description of a system using .... .hi[math]
  - Useful to help .hi[explain] and .hi[predict] behavior

--

.hi[The Canonical Example]

\begin{align*}
\text{Supply}:\hspace{.1in} P(Q_s) &= 10 + 5*Q_s\\
\text{Demand}: \hspace{.1in}P(Q_d) &= 20 - 2*Q_d
\end{align*}

--

This model allows us to make predictions about prices and quantities (from the supply & demand side), _and_ the .hi[equilibrium] price and quantity


---
# What is a model?

What are the .hi-blue[pros] of models?
  - Allows for us to be very precise with our language
  - Gives us the ability to .hi[predict] the various aspects of the economy
  - Can shed insight on .hi[mechanisms] through which processes interact
  
--
  
What are the .hi-red[cons] of models?
- They require assumptions
- .hi[Almost] all assumptions are wrong
  
--

But not all wrong assumptions are useless
    
--

The ability of the model to .hi[predict data] and .hi-blue[understand mechanisms] determines how useful it is

---
# What is a model?


Did we make assumptions our supply/demand model? .hi.purple[Discuss]


\begin{align*}
\text{Supply}:\hspace{.1in} P(Q_s) &= 10 + 5*Q_s\\
\text{Demand}: \hspace{.1in}P(Q_d) &= 20 - 2*Q_d
\end{align*}

--

1. .hi[Marginal values] are diminishing and .hi[marginal costs] are increasing <sup>.hi[†]</sup>

  - Generates downward demand and upward supply
  
.footnote[.hi[†]: Marginal = Adding one additional unit]
  
2. Demand and Supply are .hi[linear]

3. Demand and Supply are .hi[deterministic]

--

Are these reasonable? Can your behavior be explained by a simple .hi[linear function]?

--

.center[.hi[Does a function even exist?]]

---
class: inverse, middle
# Planning

.pull-left[
.hi-white[Next Class]:

- EC201 Review

- 5 Axioms of Urban Economics]

--

.pull-right[
.hi-white[Reading]:

- Get the book ASAP!!

- Read the introduction]

---

#Table of Contents

.col-left[
###Admin
.smallest[
1. [Schedule](#schedule)
1. [Syllabus](#syllabus)
]


###Intro to Urban Economics
.smallest[

1. [What is Urban Economics?](#what_is_it)

1. [What is a city?](#what_is_city)
  
1. [What is a model and why are they useful?](#what_is_model)

]
]


---
exclude: true
```{r decktape, eval = FALSE}
p_load(pagedown)
pagedown::chrome_print(here::here("001-intro","lecture_one.html"))
```





